Gradual Soundness: Lessons from Static Python

### 2022-01-24 pj-7.1/paper.tex

%% TODO before submission
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
%%   trying to avoid nonstatic? staticifying?
%% - terms: sound generic / concrete ... aren't people going to forget concrete? better keep as last resort
%% - perhaps a compatibility story, programmers can work on Python 3.9 until Cinder catches up b/c source code is the same

@@

\newcommand{\spapp}[2]{#1\,(#2)}
\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\spteval}{\typefont{S}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptX}{\typefont{X}} %% undefined type for an early example
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptstr}{\typefont{Str}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptflout}{\typefont{Flout}}
\newcommand{\sptdyn}{\typefont{Dyn}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptset}[1]{\paramtype{\typefont{Set}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptrawchkdict}{\typefont{ChkDict}} %% not a real SP type, but useful for TeX
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
\newcommand{\sptenv}{\typefont{Env}}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{\spapp{#1}{#2}\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\trule}[1]{\textsc{#1}}

@@

  \paragraph{Inquiry}
  % What problem or question does the paper
  % address? How has this problem or question been
  % addressed by others (if at all)?
  This paper describes \SP{}, a language developed by engineers at Instagram
  that has proven itself sound, fast, and reasonably expressive in production.
  \SP{}'s approach to gradual types is essentially a programmer-tunable combination of
  the \emph{concrete}\/ and \emph{transient}\/ approaches from the literature.
  Concrete types provide full soundness and low performance overhead, but impose
  nonlocal constraints.
  Transient types are sound in a shallow sense and easier to use; they help
  to bridge the gap between untyped code and typed concrete code.

@@

  \paragraph{Knowledge}
  % What new facts were uncovered? If the
  % research was not results oriented, what new
  % capabilities are enabled by the work?
  Our main finding is that the \emph{gradual soundness}\/ that
  arises from a mix of concrete and transient types is an effective
  way to lower the maintenance cost of the concrete approach.
  We also find that method-based JIT technology can eliminate the
  costs of the transient approach.
  On a more technical level, this paper describes two contributions: a model of
  \SP{} and a performance evaluation of \SP{}. The process of formalization
  found several errors in the implementation, including fatal errors.

@@

Unfortunately, the high run-time cost of sound types has split
the gradual typing community.
Industry teams have developed innovative type systems that accommodate
untyped designs, but are unsound~\cite{bat-ecoop-2014,rch-popl-2012,cvgrl-oopsla-2017}.
These systems intentionally check nothing at run-time when untyped values enter
typed code.
Academic teams have primarily focused on the theory of sound
gradual types, formulating correctness properties and studying ever-more-descriptive
types~(e.g.,~\cite{sgt-jfp-2016,clps-popl-2019,nla-popl-2019,mgt-oopsla-2021})
A few academics have studied the cost of run-time checks
in detail~\cite{tfgnvf-popl-2016,gtnffvf-jfp-2019}
and proposed implementation methods~\cite{kas-pldi-2019,bbst-oopsla-2017},
compiler technology~\cite{vsc-dls-2019,fgsfs-oopsla-2018},
and even weakened semantics~\cite{glfd-pj-2022,vss-popl-2017,gi-scp-2020},
but these efforts have not yet decisively closed the performance gap.
The most promising attempt is the \emph{concrete} semantics
for gradual types~\cite{clzv-ecoop-2018,mt-oopsla-2017,wzlov-popl-2010}, but it imposes nonlocal
constraints on untyped code.
In particular, concrete-typed client code is incompatible with
values created by untyped code.

@@

This paper reports an exception to the rule among industry-made gradual type systems.
The \SP{} team at Instagram has developed a \emph{sound} type system for a subset of
Python along with a runtime system that uses soundness to drive optimizations.
The language is staged to let programmers choose between easy migrations and
full-strength optimizations, a design that we call \emph{gradual soundness}.
To a first approximation, there are three main stages:
\begin{enumerate}
  \item
    \SP{} contains a full language of \emph{shallow types}\/ that describe idiomatic
    Python code at a coarse granularity.
    Adapting an untyped module to use shallow types requires at most a few local code
    changes.
  \item
    A small set of \emph{concrete types}\/ describe generic data structures
    that guarantee the types of their elements.
    If programmers modify their code to build concrete structures instead of Python
    ones (a potentially nonlocal change), then \SP{} can speed up reads from these
    structures.
  \item
    Additional types and flags to enable further optimizations.
    These include \emph{primitive types}\/ that enable unboxed arithmetic
    and \emph{strict module semantics}\/ that makes modules immutable.
\end{enumerate}

@@

f({"A": 1})
  \end{lstlisting}
\end{minipage}\begin{minipage}{0.08\columnwidth}\(~~\rightarrow\!\!\)\end{minipage}\begin{minipage}{0.32\columnwidth}\noindent\begin{lstlisting}
from __static__ import PyDict
def f(x: PyDict[Str, Int]):
  return x["A"]

f({"A": 1})
  \end{lstlisting}
  \end{minipage}\begin{minipage}{0.08\columnwidth}\(~~\rightarrow\!\!\)\end{minipage}\begin{minipage}{0.34\columnwidth}\noindent\begin{lstlisting}
from __static__ import ChkDict
def f(x: ChkDict[Str, Int]):
  return x["A"]

f(chkdict[Str, Int]({"A": 1}))
  \end{lstlisting}
  \end{minipage}

@@

%% TODO confirm with C&D 
%%  KC points out that variables and types are not too different so issue 50 may apply https://github.com/facebookincubator/cinder/issues/50
%%  perhaps every type should have an import; that's not a burden for Pyre users
In addition to serving as a default, the dynamic type also replaces any undefined types.
For example, suppose that the type $\sptflout{}$ is undefined in \SP{}
(i.e., it is a misspelling of $\sptfloat$) and that a
program contains the annotation $\sptchkdict{\sptstr}{\sptflout}$, which describes a
dictionary with integer keys and $\sptflout$-typed values.
\SP{} interprets this annotation as $\sptchkdict{\sptint}{\sptdyn}$ and allows any sort
of values inside the dictionary at runtime.

@@

\begin{figure}
  \begin{subfigure}[t]{0.48\columnwidth}
    \begin{lstlisting}
class A:
  def m(self) -> Int:
    return 0

class B(A):
  def m(self):
    # Error: Dyn cannot override Int
    return 0
  \end{lstlisting}
    \caption{SGG Violation: removing a type in the subclass \code{B} raises a static error}
    \label{f:gg-failure-stat}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\columnwidth}
  \begin{lstlisting}
from __static__ import ChkDict

def f(x: ChkDict[Str, Dyn]):
  return x["A"]

d = ChkDict[Str, Int]({"A": 1})
f(d)
# Error: f expected ChkDict[Str, Dyn]
  \end{lstlisting}
    \caption{DGG Violation: removing part of the type for the parameter \code{x} raises a dynamic error}
    \label{f:gg-failure-dyn}
  \end{subfigure}
  \caption{\SP{} provides neither the static (SGG) nor the dynamic (DGG) gradual guarantees (assuming a standard type-precision relation)}
  \label{fig:gg-failure}
\end{figure}

@@

Parameterized types (generics) are more difficult to enforce.
For example, suppose that a typed function expects inputs that match
the type \code{Dict[Int,Int]}; that is, dictionaries with integer keys and
integer values.
A constant-time check for dictionaries is not enough to protect the typed
function because it fails to validate the elements inside a dictionary.
The \SP{} solution is to provide two incompatible versions of every parameterized type,
both of which support efficient run-time checks.\footnote{Not all parameterized \SP{} types have both shallow
and concrete versions at present. In particular, concrete user-defined classes have yet to gain concrete support.}
For dictionaries, \SP{} provides the types $\sptrawpydict$ and $\sptrawchkdict$:
\begin{enumerate}
  \item
    The \emph{shallow} type $\sptrawpydict$ gets enforced with a constant-time check
    for dict values that ignores the elements inside a dict.
    Consequently, a parameterized surface type such as $\sptpydict{\sptint}{\sptint}$
    is merely a promise of dict values at run-time; it does not promise anything about
    dict keys and values.
  \item
    The \emph{concrete} type $\sptchkdict{K}{V}$ guarantees well-typed
    keys and values, but is incompatible with built-in Python dictionaries.
    Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
\end{enumerate}
%
These options let programmers decide whether to edit their code to support concrete types
or leave it as-is with shallow types.
Often, the shallow types are more attractive because concrete types impose
nonlocal changes.
If one type changes to concrete, then several changes may need to happen:
all values that reach this type must be initialized as concrete,
all typed clients of these values must expect concrete values,
all values that reach those clients must be concrete,
and so on and so forth.

@@

\subsection{Strict Modules}
\label{s:strict-mod}

Python modules are mutable collections of names.
Names can be added, deleted, and updated at runtime.
\SP{} accommodates these behaviors in the same manner as standard Python, which
necessarily limits the benefits of module types.
As an alternative, \SP{} programmers may opt in to a \emph{strict module} semantics.
Strict modules are immutable, and thus enable type-based compiler optimizations.

@@

@@

For example, the type $\sptchkdict{K}{V}$ describes a concrete dictionary
with keys of type $K$ and values of type $V$.
The implementation has three main components:
\begin{itemize}
  \item
    The constructor requires two type parameters and a Python dictionary ($\sptrawpydict$).
    It checks that the elements of the dictionary match the key and value types.
  \item
    All write operations check their input.
    Every operation that mutates or extends a checked dictionary must
    validate any untyped arguments that it receives.
  \item
    The dictionary value stores a compound type tag 
    (e.g., one token representing a three-part type $\sptchkdict{K}{V}$)
    to support casts from the dynamic type.
    When a concrete dictionary enters typed code from an untyped context,
    its tag must match the expected type exactly.
    For example, type $\sptchkdict{K}{V}$ matches the type
    $\sptchkdict{\sptstr}{\sptdyn}$ only if $K$ is equal to $\sptstr$
    and $V$ is equal to $\sptdyn$.
\end{itemize}
%
In general, other checked datatypes have the same three components:
a constructor, checked update functions, and a tag.

@@

In Python, the syntax for expressions overlaps with the syntax of types.
The Python name \code{None}, for example, is both an expression and a type.
By contrast, \cref{f:surface-types} keeps the two syntaxes distinct.
Expressions use lowercase names (e.g., \code{none}) and types uses capitalized
names (\code{None}).

@@

  \bigskip
  \(
    \mftypeF{\sptype_0}
    \mfeq
    \left\{\begin{array}{ll}
      \sptrawpydict & \mbox{if $\sptype_0 = \sptpydict{\sptype}{\sptype}$}
    \\
      \sptoptional{\mftypeF{\sptype_1}} & \mbox{if $\sptype_0 = \sptoptional{\sptype_1}$}
    \\
      \sptdyn & \mbox{if $\sptype_0 = \sptunion{\sptype, \ldots}$ and $\sptype_0 \neq \sptoptional{\sptype}$}
    \\
      \sptchkdict{\mftypeF{\sptype_1}}{\mftypeF{\sptype_2}} & \mbox{if $\sptype_0 = \sptchkdict{\sptype_1}{\sptype_2}$}
    \\
      \sptype_0 & \mbox{otherwise}
    \end{array}\right.
  \)

@@

    %% TODO show dict read rules?
%    \inferrule*[left=PD-Ref]{
%      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
%      \\\\
%      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
%    }{
%      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptdyn
%    }
%
%    \inferrule*[left=CD-Ref]{
%      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
%      \\\\
%      \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
%    }{
%      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \spteval_1
%    }

    \inferrule*[lab=PD-Set]{
      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
      \\\\
      \sptenv_0 \wtexpr \spexpr_2 : \sptobject
    }{
      \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
    }

@@

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle.
These include primitive types~\cref{s:c-types} strict modules~\cref{s:strict-mod},
and the boundary to untyped Python~(\cref{s:py-boundary}).
%% Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
%% are also omitted.
Other Python features are left out because their \SP{} semantics is identical
to standard Python~(\cref{s:py-boundary}).
Our model also does not cover the optimizer~(\cref{s:optimize}) because
a formal account of type-directed optimization would require substantial additional
components: a model of the \SP{} bytecode and a faithful rendering of its
transformations.

@@

There are two case in which \SP{} types impose constraints on untyped Python code.
First, concrete generic types (e.g., $\sptrawchkdict$)
reject inputs that were not created via a checked contructor.
If a client of a Python module decides to impose concrete types, the Python module
may need to change.
Second, the strict setting for modules prevents updates to module-level variables.
If existing code mutates a binding in a strict module, a run-time error will
occur.

@@

Part of this goal is met by inserting casts only in positions where the dynamic
type flows into a static type.
The materialize rule in the model illustrates this policy.
Typed functions, however, are compiled with a \bcinst{CHECK\_ARGS} instruction
that defensively casts all arguments.
By convention, this instruction always appears on the first line of a typed
function body.
The optimizer uses this convention to skip argument checks by jumping past them
when it is safe to do so; namely, whenever a typed function or method gets
called in a typed context.

@@

  \centering
  \begin{tabular}{lrr}
    Name               & Typed / Python & Refined / Python \\\midrule
    \bmname{deltablue} &          TBD &            TBD \\
    \bmname{nbody}     &          TBD &            TBD \\
    \bmname{richards}  &         0.53 &           0.22 \\
    \bmname{fannkuch}  &         1.03 &           0.46
    %% fannkuch typedbasic2 = 0.90
  \end{tabular}
\end{table}

@@

\begin{itemize}
  \item TBD need categories, expect about 40 of these.
\end{itemize}

>>>>

%% TODO before submission
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
%%   trying to avoid nonstatic? staticifying?
%% - perhaps a compatibility story, programmers can work on Python 3.9 until Cinder catches up b/c source code is the same
%% - SP doc https://docs.google.com/document/d/1-xzpBuiDYnpeGrXy61R_JCy7-iJ2v6F5DcoB4ICHEfw/edit#
%% -

@@

\newcommand{\spapp}[2]{#1\,(#2)}
\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\spteval}{\typefont{S}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptX}{\typefont{X}} %% undefined type for an early example
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptstr}{\typefont{Str}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptflout}{\typefont{Flout}}
\newcommand{\sptdyn}{\typefont{Dyn}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptset}[1]{\paramtype{\typefont{Set}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{Dict}}
\newcommand{\sptrawchkdict}{\typefont{CheckedDict}} %% not a real SP type, but useful for TeX
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\sptrawchkdict}{#1, #2}}
\newcommand{\sptenv}{\typefont{Env}}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{\spapp{#1}{#2}\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\trule}[1]{\textsc{#1}}

@@

  \paragraph{Inquiry}
  % What problem or question does the paper
  % address? How has this problem or question been
  % addressed by others (if at all)?
  This paper describes \SP{}, a language developed by engineers at Instagram
  that has proven itself sound, fast, and reasonably expressive in production.
  \SP{}'s approach to gradual types is essentially a programmer-tunable combination of
  the \emph{concrete\/} and \emph{transient\/} approaches from the literature.
  Concrete types provide full soundness and low performance overhead, but impose
  nonlocal constraints.
  Transient types are sound in a shallow sense and easier to use; they help
  to bridge the gap between untyped code and typed concrete code.

@@

  \paragraph{Knowledge}
  % What new facts were uncovered? If the
  % research was not results oriented, what new
  % capabilities are enabled by the work?
  Our main finding is that the \emph{gradual soundness\/} that
  arises from a mix of concrete and transient types is an effective
  way to lower the maintenance cost of the concrete approach.
  We also find that method-based JIT technology can eliminate the
  costs of the transient approach.
  On a more technical level, this paper describes two contributions: a model of
  \SP{} and a performance evaluation of \SP{}. The process of formalization
  found several errors in the implementation, including fatal errors.

@@

Unfortunately, the high run-time cost of sound types has split
the gradual typing community.
Industry teams have developed innovative type systems that accommodate
untyped designs, but are unsound~\cite{bat-ecoop-2014,rch-popl-2012,cvgrl-oopsla-2017}.
These systems intentionally check nothing at run-time when untyped values enter
typed code.
Academic teams have primarily focused on the theory of sound
gradual types, formulating correctness properties and studying ever-more-descriptive
types~(e.g.,~\cite{sgt-jfp-2016,clps-popl-2019,nla-popl-2019,mgt-oopsla-2021})
A few academics have studied the cost of run-time checks
in detail~\cite{tfgnvf-popl-2016,gtnffvf-jfp-2019}
and proposed implementation methods~\cite{kas-pldi-2019,bbst-oopsla-2017},
compiler technology~\cite{vsc-dls-2019,fgsfs-oopsla-2018},
and even weakened semantics~\cite{glfd-pj-2022,vss-popl-2017,gi-scp-2020},
but these efforts have not yet decisively closed the performance gap.
The most promising attempt is the \emph{concrete\/} semantics
for gradual types~\cite{clzv-ecoop-2018,mt-oopsla-2017,wzlov-popl-2010}, but it imposes nonlocal
constraints on untyped code.
In particular, concrete-typed client code is incompatible with
values created by untyped code.

@@

This paper reports an exception to the rule among industry-made gradual type systems.
The \SP{} team at Instagram has developed a \emph{sound} type system for a subset of
Python along with a runtime system that uses soundness to drive optimizations.
The language is staged to let programmers choose between easy migrations and
full-strength optimizations, a design that we call \emph{gradual soundness}.
To a first approximation, there are three main stages:
\begin{enumerate}
  \item
    A full language of \emph{shallow types\/} that describe idiomatic
    Python code at a coarse granularity.
    Adapting an untyped module to use shallow types requires at most a few local code
    changes.
  \item
    A second level of \emph{concrete types\/} describes generic data structures
    that guarantee the types of their elements.
    If programmers modify their code to build concrete structures instead of Python
    ones (a potentially nonlocal change), then \SP{} can perform deeper optimizations.
  \item
    Additional progressive types that enable further optimizations.
    In particular, \emph{primitive types\/} enable unboxed arithmetic.
\end{enumerate}

@@

f({"A": 1})
  \end{lstlisting}
\end{minipage}\begin{minipage}{0.07\columnwidth}\(~~\rightarrow\!\!\!\!\)\end{minipage}\begin{minipage}{0.32\columnwidth}\noindent\begin{lstlisting}
from typing import Dict
def f(x: Dict[str, int]):
  return x["A"]

f({"A": 1})
  \end{lstlisting}
  \end{minipage}\begin{minipage}{0.07\columnwidth}\(~~\rightarrow\!\!\!\!\)\end{minipage}\begin{minipage}{0.38\columnwidth}\noindent\begin{lstlisting}
from __static__ import CheckedDict
def f(x: CheckedDict[str, int]):
  return x["A"]

f(CheckedDict[str, int]({"A": 1}))
  \end{lstlisting}
  \end{minipage}

@@

%% TODO avoid typo ... shorten?
In addition to serving as a default, the dynamic type also replaces any undefined types.
For example, suppose that the type $\sptflout{}$ is undefined in \SP{}
(i.e., it is a misspelling of $\sptfloat$) and that a
program contains the annotation $\sptchkdict{\sptstr}{\sptflout}$, which describes a
dictionary with integer keys and $\sptflout$-typed values.
\SP{} interprets this annotation as $\sptchkdict{\sptint}{\sptdyn}$ and allows any sort
of values inside the dictionary at runtime.

@@

\begin{figure}
  \begin{subfigure}[t]{0.48\columnwidth}
    \begin{lstlisting}
class A:
  def m(self) -> int:
    return 0

class B(A):
  def m(self):
    # Error: dynamic cannot override int
    return 0
  \end{lstlisting}
    \caption{SGG Violation: removing a type in the subclass \code{B} raises a static error}
    \label{f:gg-failure-stat}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\columnwidth}
  \begin{lstlisting}
from __static__ import CheckedDict

def f(x: CheckedDict[str, dynamic]):
  return x["A"]

d = CheckedDict[str, int]({"A": 1})
f(d)
# Error: f expected CheckedDict[str, dynamic]
  \end{lstlisting}
    \caption{DGG Violation: removing part of the type for the parameter \code{x} raises a dynamic error}
    \label{f:gg-failure-dyn}
  \end{subfigure}
  \caption{\SP{} provides neither the static (SGG) nor the dynamic (DGG) gradual guarantees (assuming a standard type-precision relation)}
  \label{fig:gg-failure}
\end{figure}

@@

Parameterized types (generics) are more difficult to enforce.
For example, suppose that a typed function expects inputs that match
the type \code{Dict[int,int]}; that is, dictionaries with integer keys and
integer values.
A constant-time check for dictionaries is not enough to protect the typed
function because it fails to validate the elements inside a dictionary.
The \SP{} solution is to provide two incompatible versions of every parameterized type,
both of which support efficient run-time checks.\footnote{Not all parameterized \SP{} types have both shallow
and concrete versions at present.
In particular, user-defined classes do not have concrete support.
The shallow check for a user-defined class, however, guarantees the types of all
non-generic fields and methods.}
For dictionaries, \SP{} provides the types $\sptrawpydict$ and $\sptrawchkdict$:
\begin{enumerate}
  \item
    The \emph{shallow} type $\sptrawpydict$ gets enforced with a constant-time check
    for dict values that ignores the elements inside a dict.
    Consequently, a parameterized surface type such as $\sptpydict{K}{V}$
    is merely a promise of dict values at run-time; it does not promise anything about
    dict keys and values.
  \item
    The \emph{concrete} type $\sptchkdict{K}{V}$ guarantees well-typed
    keys and values, but is incompatible with built-in Python dictionaries.
    Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
\end{enumerate}
%
These options let programmers decide whether to edit their code to support concrete types
or leave it as-is with shallow types.
Often, the shallow types are more attractive because concrete types impose
nonlocal changes.
If one type changes to concrete, then several changes may need to happen:
all values that reach this type must be initialized as concrete,
all typed clients of these values must expect concrete values,
all values that reach those clients must be concrete,
and so on and so forth.

@@

@@

\subsection{Behavioral Changes to Python}
\label{s:not-python}

FILL

%% Python modules are mutable collections of names.
%% Names can be added, deleted, and updated at runtime.
%% \SP{} accommodates these behaviors in the same manner as standard Python, which
%% necessarily limits the benefits of module types.
%% As an alternative, \SP{} programmers may opt in to a \emph{strict module} semantics.
%% Strict modules are immutable, and thus enable type-based compiler optimizations.

%% 1 strict modules
%% 2 .....

@@

For example, the type $\sptchkdict{K}{V}$ describes a concrete dictionary
with keys of type $K$ and values of type $V$.
The implementation has three main components:
\begin{itemize}
  \item
    The constructor requires two type parameters and a Python dictionary ($\sptrawpydict$).
    It checks that the elements of the dictionary match the key and value types.
  \item
    All write operations check their input.
    Every operation that mutates or extends a checked dictionary must
    validate any untyped arguments that it receives.
  \item
    The dictionary value stores a compound type tag 
    (e.g., one token representing a three-part type $\sptchkdict{K}{V}$)
    to support casts from the dynamic type.
    When a concrete dictionary enters typed code from an untyped context,
    its tag must match the expected type exactly.
    For example, type $\sptchkdict{K}{V}$ matches the type
    $\sptchkdict{K'}{V'}$ only if $K$ is equal to $K'$
    and $V$ is equal to $V'$.
\end{itemize}
%
In general, other checked datatypes have the same three components:
a constructor, checked update functions, and a tag.

@@

In Python, the syntax for expressions overlaps with the syntax of types.
The Python name \code{None} is both an expression and a type,
and the \SP{} names \code{chkdict} and \code{CheckedDict} are synonyms.
By contrast, \cref{f:surface-types} keeps the two syntaxes distinct.
Expressions use lowercase names (\code{none}, \code{chkdict}) and types uses capitalized
names (\code{None}, \code{CheckedDict}).

@@

  \bigskip
  \(
    \mftypeF{\sptype_0}
    \mfeq
    \left\{\begin{array}{ll}
      \sptrawpydict & \mbox{if $\sptype_0 = \sptpydict{\sptype}{\sptype}$}
    \\
      \sptoptional{\mftypeF{\sptype_1}} & \mbox{if $\sptype_0 = \sptoptional{\sptype_1}$}
    \\
      \sptdyn & \mbox{if $\sptype_0 = \sptunion{\sptype, \ldots}$ and $\sptype_0 \neq \sptoptional{\sptype}\!\!\!\!$}
    \\
      \sptchkdict{\mftypeF{\sptype_1}}{\mftypeF{\sptype_2}} & \mbox{if $\sptype_0 = \sptchkdict{\sptype_1}{\sptype_2}$}
    \\
      \sptype_0 & \mbox{otherwise}
    \end{array}\right.
  \)

@@

    %% TODO show dict read rules?
%    \inferrule*[left=D-Ref]{
%      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
%      \\\\
%      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
%    }{
%      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptdyn
%    }
%
%    \inferrule*[left=CD-Ref]{
%      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
%      \\\\
%      \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
%    }{
%      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \spteval_1
%    }

    \inferrule*[lab=D-Set]{
      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
      \\\\
      \sptenv_0 \wtexpr \spexpr_2 : \sptobject
    }{
      \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
    }

@@

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle.
These include primitive types~\cref{s:c-types}, module-level variables,
and the boundary to untyped Python~(\cref{s:py-boundary}).
%% Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
%% are also omitted.
Other Python features are left out because their \SP{} semantics is identical
to standard Python~(\cref{s:py-boundary}).
Our model also does not cover the optimizer~(\cref{s:optimize}) because
a formal account of type-directed optimization would require substantial additional
components: a model of the \SP{} bytecode and a faithful rendering of its
transformations.

@@

%% TODO FILL only 2 cases? revisit after adding "changes to python" section in 2
There are two cases in which \SP{} types impose constraints on untyped Python code.
First, concrete generic types (e.g., $\sptrawchkdict$)
reject inputs that were not created via a checked contructor.
If a client of a Python module decides to impose concrete types, the Python module
may need to change.
Second, typed modules prevent updates to module-level variables.

@@

Part of this goal is met by inserting casts only in positions where the dynamic
type flows into a static type.
The materialize rule in the model illustrates this policy, which ensures that
well-typed writes to classes and to concrete dictionaries have no cost.
Typed functions, however, are compiled with a \bcinst{CHECK\_ARGS} instruction
that defensively casts all arguments.
By convention, this instruction always appears on the first line of a typed
function body.
The optimizer uses this convention to skip argument checks by jumping past them
when it is safe to do so; namely, whenever a typed function or method gets
called in a typed context.

@@

  \centering
  \begin{tabular}{lrr}
    Name               & Typed / Python & Refined / Python \\\midrule
    \bmname{deltablue} &         FILL &           FILL \\
    \bmname{nbody}     &         FILL &           FILL \\
    \bmname{richards}  &         0.53 &           0.22 \\
    \bmname{fannkuch}  &         1.03 &           0.46
    %% fannkuch typedbasic2 = 0.90
  \end{tabular}
\end{table}

@@

\begin{itemize}
  \item FILL need categories, expect about 40 of these.
\end{itemize}
